---
title: "PA2_Project"
author: "Paul Vinod"
date: "12/09/2020"
output:
  html_document:
    df_print: paged
---

Weather Affect on Human and Economic 
==========================================

This is a report on National Storm data which would be focusing on following key objectives:
- How much harm has been provide to population?
- How much damage it has been provided on the economy?

The first question is addressed by condiering the fatalities and injuries. Whereas the second question includes crop and property damages. 

Collecting the data from online sources:
```{r}
# Loading libraries.
library(dplyr)
library(grDevices)
library(ggplot2)

# Download files
filename = "CourseProject2.csv.bz2"
if(!file.exists(filename))
{
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl,
                  destfile = filename,
                  method = "curl")

}

valDT <- read.csv(bzfile("CourseProject2.csv.bz2"),
                  header = TRUE)
x <- dim(valDT)
```
5
The data contains `x[1]` observations and `x[2]` variables. Our focus would be on the variable EVTYPE which contains various parameters. Clubing these parameters to following: "FLOOD"   "HAIL"    "HEAT"    "OTHER"   "RAIN"    "SNOW"    "STORM"   "TORNADO" "WIND"    "WINTER" 

```{r}
valDT$EVENT <- "OTHER"
valDT$EVENT[grep("RAIN", valDT$EVTYPE, ignore.case = TRUE)] <- "RAIN"
valDT$EVENT[grep("HAIL", valDT$EVTYPE, ignore.case = TRUE)] <- "HAIL"
valDT$EVENT[grep("TORNADO", valDT$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
valDT$EVENT[grep("HEAT", valDT$EVTYPE, ignore.case = TRUE)] <- "HEAT"
valDT$EVENT[grep("SNOW", valDT$EVTYPE, ignore.case = TRUE)] <- "SNOW"
valDT$EVENT[grep("STORM", valDT$EVTYPE, ignore.case = TRUE)] <- "STORM"
valDT$EVENT[grep("WINTER", valDT$EVTYPE, ignore.case = TRUE)] <- "WINTER"
valDT$EVENT[grep("WIND", valDT$EVTYPE, ignore.case = TRUE)] <- "WIND"
valDT$EVENT[grep("FLOOD", valDT$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
```

Similary converting the property and crop damage expenses to numeric values. 
K = 10^3; M = 10^6, B = 10^9
```{r results = FALSE}
valDT$PROPDMGEXP[!grepl("K|M|B", valDT$PROPDMGEXP,ignore.case = T)] < - 0
valDT$CROPDMGEXP[!grepl("K|M|B", valDT$CROPDMGEXP,ignore.case = T)] < - 0

valDT$PROPDMGEXP[valDT$PROPDMGEXP == "B"|valDT$PROPDMGEXP == "b" ]  <- 9
valDT$PROPDMGEXP[valDT$PROPDMGEXP == "M"|valDT$PROPDMGEXP == "m" ]  <- 6
valDT$PROPDMGEXP[valDT$PROPDMGEXP == "K"|valDT$PROPDMGEXP == "k" ]  <- 3
valDT$PROPDMGEXP <- as.numeric(valDT$PROPDMGEXP)
valDT$PROPDMGEXP[is.na(valDT$PROPDMGEXP)] <- 0

valDT$CROPDMGEXP[valDT$CROPDMGEXP == "B"|valDT$CROPDMGEXP == "b" ]  <- 9
valDT$CROPDMGEXP[valDT$CROPDMGEXP == "M"|valDT$CROPDMGEXP == "m" ]  <- 6
valDT$CROPDMGEXP[valDT$CROPDMGEXP == "K"|valDT$CROPDMGEXP == "k" ]  <- 3
valDT$CROPDMGEXP <- as.numeric(valDT$CROPDMGEXP)
valDT$CROPDMGEXP[is.na(valDT$CROPDMGEXP)] <- 0
```

Calculating the overall damages caused by each of property and crop. 
```{r}
valDT$cropdamage <- valDT$CROPDMG * 10^valDT$CROPDMGEXP
valDT$propdamage <- valDT$PROPDMG * 10^valDT$PROPDMGEXP
```

Grouping these collected information by EVENT. 
```{r}
valDT_clean <- valDT %>% group_by(EVENT) %>%
               summarise(net_fatal = sum(FATALITIES),
                net_injuries = sum(INJURIES),
                net_crop_damage = sum(cropdamage),
                net_prop_damage = sum(propdamage))
print(valDT_clean)
```

Plotting the responses of **Human health and population**
```{r}
par(mfrow = c(1,2))
barplot(height = valDT_clean$net_fatal, names.arg = valDT_clean$EVENT,
        col = "red", xlab = "Events", ylab = "Fatalities", ylim = c(0,6000))

barplot(height = valDT_clean$net_injuries, names.arg = valDT_clean$EVENT,
        col = "blue",  xlab = "Events", ylab = "Injuries", ylim = c(0,95000))
title(main = "Human health and population Loss")

```
Plotting the responses of **Economy loss by crop and property damages**
```{r}
par(mfrow = c(2,1))
barplot(height = valDT_clean$net_crop_damage/10000000, names.arg = valDT_clean$EVENT,
        col = "blue", width = 2, ylim = c(0,2500))

barplot(height = valDT_clean$net_prop_damage/10000000, names.arg = valDT_clean$EVENT,col = "red", width = 2, ylim = c(0, 20000))

title(main = "Economy Loss by Crop and Property Damage")
```
