---
title: "PA2_Project"
author: "Paul Vinod"
date: "12/09/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Weather Affect on Human and Economic 
## 1. Introduction

This is a report on National Storm data which would be focusing on following key objectives:
- How much harm has been provide to population?
- How much damage it has been provided on the economy?

The first question is addressed by condiering the fatalities and injuries. Whereas the second question includes crop and property damages. 

Collecting the data from online sources:
```{r CACHE = TRUE}
# Loading libraries.
library(dplyr)
library(grDevices)
library(ggplot2)
library(tidyr)

# Download files
filename = "CourseProject2.csv.bz2"
if(!file.exists(filename))
{
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileUrl,
                  destfile = filename,
                  method = "curl")

}

valDT <- read.csv(bzfile("CourseProject2.csv.bz2"),
                  header = TRUE)
x <- dim(valDT)
```

### 1.1 Dimemnsion of Data Set
The data contains `r x[1]` observations and `r x[2]` variables. Our focus would be on the variable EVTYPE which contains various parameters.

## 2. Data Processing

The data need to be cleaned on the basis of EVTYPE and modified to the following list. 
"FLOOD"   "HAIL"    "HEAT"    "OTHER"   "RAIN"    "SNOW"    "STORM"   "TORNADO" "WIND"    "WINTER" 
```{r}
valDT$EVENT <- "OTHER"
valDT$EVENT[grep("RAIN", valDT$EVTYPE, ignore.case = TRUE)] <- "RAIN"
valDT$EVENT[grep("HAIL", valDT$EVTYPE, ignore.case = TRUE)] <- "HAIL"
valDT$EVENT[grep("TORNADO", valDT$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
valDT$EVENT[grep("HEAT", valDT$EVTYPE, ignore.case = TRUE)] <- "HEAT"
valDT$EVENT[grep("SNOW", valDT$EVTYPE, ignore.case = TRUE)] <- "SNOW"
valDT$EVENT[grep("STORM", valDT$EVTYPE, ignore.case = TRUE)] <- "STORM"
valDT$EVENT[grep("WINTER", valDT$EVTYPE, ignore.case = TRUE)] <- "WINTER"
valDT$EVENT[grep("WIND", valDT$EVTYPE, ignore.case = TRUE)] <- "WIND"
valDT$EVENT[grep("FLOOD", valDT$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
```

Similarly converting the property and crop damage expenses to numeric values. 
K = 10^3; M = 10^6, B = 10^9
```{r results = FALSE}
valDT$PROPDMGEXP[!grepl("K|M|B", valDT$PROPDMGEXP,ignore.case = T)] < - 0
valDT$CROPDMGEXP[!grepl("K|M|B", valDT$CROPDMGEXP,ignore.case = T)] < - 0

valDT$PROPDMGEXP[valDT$PROPDMGEXP == "B"|valDT$PROPDMGEXP == "b" ]  <- 9
valDT$PROPDMGEXP[valDT$PROPDMGEXP == "M"|valDT$PROPDMGEXP == "m" ]  <- 6
valDT$PROPDMGEXP[valDT$PROPDMGEXP == "K"|valDT$PROPDMGEXP == "k" ]  <- 3
valDT$PROPDMGEXP <- as.numeric(valDT$PROPDMGEXP)
valDT$PROPDMGEXP[is.na(valDT$PROPDMGEXP)] <- 0

valDT$CROPDMGEXP[valDT$CROPDMGEXP == "B"|valDT$CROPDMGEXP == "b" ]  <- 9
valDT$CROPDMGEXP[valDT$CROPDMGEXP == "M"|valDT$CROPDMGEXP == "m" ]  <- 6
valDT$CROPDMGEXP[valDT$CROPDMGEXP == "K"|valDT$CROPDMGEXP == "k" ]  <- 3
valDT$CROPDMGEXP <- as.numeric(valDT$CROPDMGEXP)
valDT$CROPDMGEXP[is.na(valDT$CROPDMGEXP)] <- 0
```

### 2.1 Calculating the Crop and Property Damage
Calculating the overall damages caused by each of property and crop. 
```{r}
valDT$cropdamage <- valDT$CROPDMG * 10^valDT$CROPDMGEXP
valDT$propdamage <- valDT$PROPDMG * 10^valDT$PROPDMGEXP
```

Grouping these collected information by EVENT. 
```{r}
valDT_pop <- valDT %>% group_by(EVENT) %>%
                summarise(net_fatal = sum(FATALITIES),
                net_injuries = sum(INJURIES)) %>%
                gather("TYPE", "Value", -EVENT)
valDT_eco <- valDT %>% group_by(EVENT) %>%
                summarise(net_crop_damage = sum(cropdamage),
                net_prop_damage = sum(propdamage)) %>%
                gather("TYPE", "Value", -EVENT)
print(valDT_pop)
print(valDT_eco)
```

## 3. Results

### 3.1Plotting the information
Plotting the information of the damages caused on the human capital by visualizing the fatalities and injuries in perspective of each EVENT.

### 3.2 Plotting of Population Effect
```{r}
plot_pop = ggplot(valDT_pop, aes(x = EVENT, y = Value, fill = TYPE)) +
       geom_bar(stat = "identity")+
       facet_wrap(TYPE~.)
print(plot_pop)
```


### 3.3 Plotting of Economic Damage
```{r}
plot_eco = ggplot(valDT_eco, aes(x = EVENT, y = Value, fill = TYPE)) +
       geom_bar(stat = "identity")+
       facet_wrap(TYPE~.)
print(plot_eco)
```

## 4. Summary
Thus with the analysis of the above two graphs we can see that the National Weather caused an immense impact on the both population and economics. Key takeaways include:
-Population health impact graph Tornado had the highest fatality.
-Economic graph the Flood caused highes impact on the property. 
-In Economic graph the effect on the crop was relatively less. 
